\textbf{Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-08-23 submission/manuscript.tex
add interpretation data and fix after mmeting 4

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our sample size from 490 to 245,
120, 60, 30 and 15 {[}Figure S4{]}. The range among the cross-validation
AUROC values within both models at lower sample sizes were much larger
than when full sample size was used to train and validate the models.
These results showed that because microbiome data has many features
(6920 OTUs), it is important to train models using appropriate sample
sizes to avoid generalizability issues.

@@

We interpreted the feature importance in the linear models (L1 and
L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights of each
OTU {[}Figure 3{]}. The OTUs that had the largest median rank and drove
the detection of SRNs across the linear models contained sequences that
were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742),
Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659,
OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus
(OTU 367) and others {[}Figure 3{]}. Among the highest ranked OTUs
across these three models, it was encouraging that many OTUs (e.g.~OTU
822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the
models.

We interpreted the feature importance in the non-linear models using
permutation importance. Whereas the absolute feature weights were
determined from the trained models, here we measured importance using
the held-out test data. Permutation importance analysis is a posthoc
explanation of the model where we randomly permuted non-correlated
features individually and groups of perfectly correlated features across
the two groups. We then calculated how much the predictive performance
of the model (i.e testing AUROC values) decreased when each OTU or group
of OTUs was randomly permuted. We ranked the OTUs based on how much the
median testing AUROC decreased when it was permuted; the OTU with the
largest decrease ranked highest. The OTUs that had the largest impact
and drove the detection of SRNs across the non-linear models contained
sequences that were most similar to members of the Lachnospiraceae (OTU
50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629),
Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58),
Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477),
Parvimonas (OTU 361), Roseburia (OTU 822) and others {[}Figure 4{]}.
Among the twenty OTUs with the largest impact across these four models,
there was only one OTU (OTU 822) that was shared among all of the
models; however, among the tree-based models, we found three OTUs (OTU
367, OTU 58 and OTU 110) to be consistently important. Similarly, in the
random forest and XGBoost models, four of the twenty most important OTUs
(OTU 361, 2, 477, 12) were shared between the models.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models {[}Figure S3{]}.
When we analyzed the L1-regularized SVM with linear kernel model using
permutation importance, the approach picked out 17 out of 20 top OTUs
(e.g.~OTU 822, OTU 1239, OTU 609) that were deemed important when we
interpreted the model using feature rankings based on weights {[}Figure
3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
regression each identified many OTUs that were selected by both
interpretation methods, 9 out of 20 and 12 out of 20 respectively.

Interestingly, for the all the models, except the poorly performing
L1-regularized SVM and decision tree, the variation in the impact of
these important OTUs was minimal. This supports the findings of others
that a classification requires numerous OTUs, rather the presence of a
single or a few of the OTUs (32).

\textbf{The computational efficiency of each ML model.}

We compared the training times of the seven ML models. As expected, the
training times increased as the complexity of a ML model and the number
of tuned hyperparameter settings increased and the linear models trained
faster than non-linear models {[}Figures S1-S2; Figure 5{]}. When we
subsetted the size of the training dataset, we observed a {[}linear,
quadratic, etc{]} relationship between the size of the dataset and the
training time for the random forest model.

@@

Overview • Establish pipeline that can be generalized to any modeling
method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
• Importance for transparency, picking the most complex model is not
necessarilty the best model • Good / better / best: Table S1 • ``best''
has multiple meanings Point 2 • Our paper is concerned with developing
models to gain greater biological insights into what's going on • Still
need biological validation and further experimentation Point 3 • These
models show great opportunity to use microbiome to make diagnoses • We
talked about classification, but others have used it in microbiome data
to do regression, mulit-class -- pipeline is the same • Others have used
this on shotgun metagenomic data and collinearity is expected to be an
even larger problem since 5000 genes are all correlated with each other
• Deployment is a different question that requires independent datasets
to test on; we currently lack decent sized parallel datasets where the
same type of sequence data and clinical data have been collected • We
also have a problem with OTU-based approaches Conclusion/Future
direction • ML is awesome and there's always new methods; our framework
gives structure to investigators wanting to try these out. We encourage
reseraches to apply multiple models to their data and report the results
• Things like neural nets are attractive and hold great promise, but we
need a ton more data to apply and they are impossible to interpret

Microbiome studies use ML models, often with a classification task, to
predict a disease but also to learn which microbes are indicators of
that disease (2--11). Achieving either of these tasks have far-reaching
impact on human health, however ML as a tool in microbiome studies is
still at its infancy. A framework is needed to develop rigorous ML
models, to identify and overcome potential pitfalls. Previous studies
generated workflows to allow ML to be widely used by the microbiome
researchers (20--23). This study sets-up standards for developing and
evaluating rigorous ML models for microbiome data {[}Table 1{]}.

We benchmarked seven ML models with different classification algorithms
to show that a clearly defined ML problem that is based on the goal of
the microbiome study should inform our model selection. Our results
showed that if the goal of a study is to learn the ecology behind a
disease and to identify microbial biomarkers, we can create ML models
that are inherently interpretable and easily trained without losing
predictive power. In terms of predictive performance, random forest
model had the best testing AUROC values compared to the other six
models. However, the second best model was L2-regularized logistic
regression with a median AUROC difference of only 0.015 compared to
random forest. While random forest took 83.2 hours to train,
L2-regularized logistic regression trained in 12 minutes. In terms of
interpretability, random forest was a more complex ML model and could
only be explained using methods such as permutation importance. On the
other hand, L2-regularized logistic regression was easier to interpret
by ranking absolute feature weights of the trained model.

Even with interpretable models such as L2-regularized logistic
regression, there are potential pitfalls when it comes to identifying
biomarkers of a disease. As domain experts, we know that
human-associated microbial communities have complex correlation
structures that creates collinearity in the dataset. Collinearity is a
severe problem and needs to be addressed to reliably interpret ML models
(33). In this study we used two different methods to interpret our
linear models; ranking each OTU by (1) their absolute weights in the
trained models and (2) their impact on the predictive performance based
on permutation importance. We observed differences in the OTU rankings
between the two interpretation methods due to collinearity in the
dataset. To avoid misinterpreting the models, once we identify the most
important microbes, we should check for their relationships with other
microbes as well. These relationships will help us generate new
hypotheses about the ecology of the disease. These hypotheses needs to
be tested with follow-up experiments to identify the true biomarkers of
a disease.

In this study, we also established a rigorous ML pipeline to use 16S
rRNA sequence counts to predict a binary health outcome. First of all,
we used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the difference between
cross-validation and test performance is low, this suggest the models
are not overfit and that they will perform similar with similar data. In
all the models, the difference between median cross-validation and
testing AUROC values did not exceed 0.021 which suggests that these
models are generalizable and can be used to test similar new data.
Furthermore, we performed the initial 80\%-20\% random datasplit 100
times in our ML pipeline. Depending on how the data is split, there is
the chance of being overoptimistic about the predictive performance of a
model. We showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average
between different data-splits. The randomization and resampling of the
initial data-split to create a held-out test set is a crucial step in
the ML pipeline to develop robust ML models and to report reliable
performance metrics. Additionally, we performed a full grid search for
hyperparameter settings when training our ML models. Default
hyperparameter settings in previously developed ML packages in R,
Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default. And finally, we used
the AUROC metric in our study to evaluate the predictive performance of
the ML models. AUROC is always random at the value 0.5 and is a robust
metric when a dataset is imbalanced.

We used a balanced CRC dataset to develop ML models with a binary
classification task. We did not evaluate multicategory classification
methods or regression analyses to predict non-binary outcomes. However,
the principles highlighted throughout this study {[}Table 1{]} apply to
all ML modeling tasks with microbiome data. The models we built were
generalizable despite the high number of features microbiome datasets
usually have. The generalization performance of ML models depends on
sample size. The more complex the model, the more data it will need. Our
dataset had 490 samples, however microbiome studies that have smaller
sample sizes would benefit from using less complex models. Our analysis
was limited to shallow learning methods and did not explore deep
learning methods such as neural networks. Microbiome datasets often
suffer from having high dimensionality but low sample sizes which makes
deep learning models prone to overfitting. There are studies that
address overcoming these challenges in biomedical datasets (11, 34, 35),
however sudies that estalish frameworks with microbiome data are
lacking. This would be an interesting direction for future work in
microbiome studies.

>>>>

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new dataset used to generate a model. In the example
of L1-regularized SVM with linear kernel {[}Figure S1{]}, the model
showed large variability between different regularization coefficients
(C) and was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
range among the cross-validation AUROC values within both models at
lower sample sizes were much larger than when the full collection of
samples was used to train and validate the models. These results showed
that because the microbiome data had many features (6920 OTUs), it was
important to train the models using appropriate sample sizes to avoid
problems with generalizability and over-optimism. Furthermore, it was
encouraging that even for a small number of samples, the 95\% confidence
interval included the median AUROC values for the larger subsetted
datasets.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed at the signs of feature weights to
determine whether an OTU is associated with classifying a subject as
being healthy or having an SRN - negative sign indicated being healthy
and positive sign indicated having SRN. Among the highest ranked OTUs
across these three models, it was encouraging that many of the same OTUs
(e.g.~OTU 50, 426, 609, 822, 1239) were important across all of the
models. The benefit of this approach is that the results of the analysis
are based on the trained model parameters and provide information
regarding the sign and magnitude of the impact of each OTU. However,
this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance. Whereas the absolute feature weights were determined from
the trained models, here we measured importance using the held-out test
data. Permutation importance analysis is a posthoc explanation of the
model where we randomly permuted non-correlated features individually
and groups of perfectly correlated features across the two groups in the
held-out test data. We then calculated how much the predictive
performance of the model (i.e testing AUROC values) decreased when each
OTU or group of OTUs was randomly permuted. We ranked the OTUs based on
how much the median testing AUROC decreased when it was permuted; the
OTU with the largest decrease ranked highest {[}Figure 4{]}. Among the
twenty OTUs with the largest impact for each of these models, there was
only one OTU (OTU 822) that was shared among all of the models; however,
we found three OTUs (OTU 58, 110, 367) that were important in each of
the tree-based models. Similarly, the random forest and XGBoost models,
shared four of the most important OTUs (OTU 2, 12, 361, 477).
Permutation analysis results also revealed that with the exception of
the decision tree model, removal of any individual OTU had a minimal
impact on model performance. For example, if OTU 367 were permuted
across the diagnoses from the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if we permuted the same OTU
from the random forest model, the AUROC only dropped from 0.695 to
0.6799. Effectively, the complexity of the communities was more fully
represented in the better performing models(22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. While permutation analysis allowed us to gauge the
importance of an OTU, the analysis is post-hoc (i.e.~done using the test
data) and these results did not allow us to directly interrogate the
models to know whether an OTU is associated with classifying a subject
as being healthy or having an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S3{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S3{]}, 17 out of the 20 top OTUs
(e.g.~OTU 609, 822, 1239) were deemed important by both approaches.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
approaches. These results indicate that both approaches are consistent
in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of tuned
hyperparameter settings. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
size of the training dataset, we observed a {[}linear, quadratic, etc{]}
relationship between the size of the dataset and the training time for
the random forest model.

@@

Overview • Establish pipeline that can be generalized to any modeling
method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
• Importance for transparency, picking the most complex model is not
necessarilty the best model • Good / better / best: Table S1 • ``best''
has multiple meanings Point 2 • Our paper is concerned with developing
models to gain greater biological insights into what's going on • Still
need biological validation and further experimentation

Point 3 • These models show great opportunity to use microbiome to make
diagnoses • We talked about classification, but others have used it in
microbiome data to do regression, mulit-class -- pipeline is the same •
Others have used this on shotgun metagenomic data and collinearity is
expected to be an even larger problem since 5000 genes are all
correlated with each other • Deployment is a different question that
requires independent datasets to test on; we currently lack decent sized
parallel datasets where the same type of sequence data and clinical data
have been collected • We also have a problem with OTU-based approaches
Conclusion/Future direction • ML is awesome and there's always new
methods; our framework gives structure to investigators wanting to try
these out. We encourage reseraches to apply multiple models to their
data and report the results • Things like neural nets are attractive and
hold great promise, but we need a ton more data to apply and they are
impossible to interpret

There is a growing awareness that many human diseases and environmental
processes are not driven by single organism, but are the product of
multiple bacterial populations. Whereas traditional statistical
approaches are useful for identifying those cases where a single
organism can drive a process, ML methods offer the ability to use many
features of these communities to identify those cases. If it is possible
to classify communities reliably, then it is hoped that we could
identify those populations that are responsible for the classification.
Because application of ML in microbiome studies is still in its infancy,
we developed a framework to train rigorous, transparent and reproducible
ML models {[}Table 1{]}.

To test this framework, we benchmarked seven ML models to show that a
clearly defined ML problem that is based on the goal of the microbiome
study should inform our model selection. Our results showed that if the
goal of a study is to learn the ecology behind a disease and to identify
microbial biomarkers, we can create ML models that are inherently
interpretable and easily trained without losing predictive power. In
terms of predictive performance, random forest model had the best
testing AUROC values compared to the other six models. However, the
second best model was L2-regularized logistic regression with a median
AUROC difference of only 0.015 compared to random forest. While random
forest took 83.2 hours to train, L2-regularized logistic regression
trained in 12 minutes. In terms of interpretability, random forest was a
more complex ML model and could only be explained using post-hoc methods
such as permutation importance. On the other hand, L2-regularized
logistic regression was easier to interpret by ranking absolute feature
weights of the trained model.

Even with interpretable models such as L2-regularized logistic
regression, there are potential pitfalls when it comes to identifying
biomarkers of a disease. Human-associated microbial communities have
complex correlation structures. This can hinder our ability to reliably
interpret models (33). In this study we used two different
interpretation methods; ranking each OTU by (1) their absolute weights
in the trained models and (2) their impact on the predictive performance
based on permutation importance. The feature weights of correlated OTUs
are influenced by and dependent on one another, which makes it harder to
identify the importance of unique OTUs. To avoid misinterpreting the
models, once we identify highly ranked OTUs, we should check for their
relationships with other OTUs as well. These relationships will help us
generate new hypotheses about the ecology of the disease and test with
follow-up experiments. On the other hand, when we used permutation
importance, we grouped correlated OTUs to determine their impact as a
group. In this study, we grouped OTUs that had a perfect correlation
with each other however, we can reduce the correlation threshold to
further investigate the relationships among features to identify the
true underlying factors when making a classification.

In this study, we established a pipeline that can be generalized to any
modeling method that uses 16S rRNA sequence counts to predict a binary
health outcome. We used a held-out test set and we performed the initial
80\%-20\% random datasplit to create the held-out test set 100 times in
our ML pipeline. The models we built with this pipeline were
generalizable despite the high number of features microbiome datasets
usually have.

We used a CRC dataset to develop ML models with a binary classification
task. We did not evaluate multicategory classification methods or
regression analyses to predict non-binary outcomes. However, the
principles highlighted throughout this study {[}Table 2{]} apply to all
ML modeling tasks with microbiome data. Our analysis was also limited to
shallow learning methods and did not explore deep learning methods such
as neural networks. Microbiome datasets often suffer from having high
dimensionality but low sample sizes which makes deep learning models
prone to overfitting.

There are studies that address overcoming these challenges in biomedical
datasets (11, 34, 35), however sudies that estalish frameworks with
microbiome data are lacking. This would be an interesting direction for
future work in microbiome studies.
